Machine Learning Tools for Open Cluster Characterization with Gaia DR2 Data

### 2021-01-29 paper/paper.tex
paper: Update authors
paper: Add Aims and Method

\documentclass[11pt,a4paper,USenglish,twocolumn]{article}
\usepackage{unir_paper}

%---------------------------
%título del trabajo y autor
%---------------------------
\title{Machine Learning Tools for Open Cluster Characterization with Gaia DR2 Data}
\author{Carlos David Álvaro Yunta}
\date{24th of January, 2021}

@@

Objetivo general, objetivos específicos y metodología de trabajo aplicada.

>>>>

\documentclass[11pt,a4paper,USenglish,twocolumn]{article}
\usepackage{unir_paper}
\usepackage{subcaption}

%---------------------------
%título del trabajo y autor
%---------------------------
\title{Machine Learning Tools for Open Cluster Characterization with Gaia DR2 Data}
\author{C. D. Álvaro, J. Álvaro, C. A. Guzmán}
\date{24th of January, 2021}

@@

Our aim is to \emph{build an unsupervised clustering model for open cluster characterization}.
The model will be \emph{non-supervised and non-parameterized} in order to fit a wide range
of clusters without the need for fine-tuning a high number of hyperparameters.

\subsection{DOWNLOAD}

We start by selecting a region from the OpenClust catalogue and downloading it from
Gaia DR2 database into a docker database. This allows us to repeat the analysis without
having to download the data every time. The radius of the downloaded region from Gaia
is 1.5 times larger than the one registered in OpenClust. This way we ensure to include
several stars that do not belong to the open cluster.

\subsection{FEATURE SELECTION}
\label{sec:feature_selection}

As mentioned before, we want our model to characterize open clusters by looking at
their dynamic properties. Also, we want to maintain as simple as possible our
clustering model in order to save computing resources.

Proper motion in right ascension and declination seems like a natural choice since,
as we know, stars belonging to the same OC share a common motion vector.

Parallax is another important feature. It lets us know how far stars are from us.
In addition, since all stars within an open cluster were born from the same dust cloud,
they must all have similar parallax.

However, we are not going to use these raw features.
Instead, we are taking a combination of them.

First, we correct proper motion in right ascension and declination by dividing them by
the parallax. That way, we normalize these quantities and help our clustering models
to improve their performance.

The modulus of the proper motion is another computed property that we are considering.
We use it to relate both features and therefore, to force our model to keep them tight.

\begin{figure}
  \includegraphics[width=\columnwidth]{../figures/melotte_22/features_melotte_22.png}
  \caption{Pairwise relationships among variables using Melotte 22 data}
  \label{fig:features_melotte_22}
\end{figure}

Figure \ref{fig:features_melotte_22} shows a pairwise relationship among some features
available in the dataset for Melotte 22 data.

\subsection{SOFT CLUSTERING}

Our first approach to find the open cluster is using the K-Means algorithm.
Since we are looking for a single cluster, it seems reasonable to use a clustering
algorithm and set it to find two clusters. One for the desired OC and another which
contains stars that do not belong to the open cluster.
However, this idea is not completely right.

\begin{figure}
  \centering
  \begin{subfigure}{\columnwidth}
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans/kmeans_n2_pm_melotte_22.png}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans/kmeans_n5_pm_melotte_22.png}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans/kmeans_n8_pm_melotte_22.png}
    \end{subfigure}
  \end{subfigure}
  \medskip
  \begin{subfigure}{\columnwidth}
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans/kmeans_n2_parallax_melotte_22.png}
      \caption{N = 2}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans/kmeans_n5_parallax_melotte_22.png}
      \caption{N = 5}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans/kmeans_n8_parallax_melotte_22.png}
      \caption{N = 8}
    \end{subfigure}
  \end{subfigure}
  \caption{K-Means comparisons with Melotte 22}
  \label{fig:kmeans_comparisons_melotte_22}
\end{figure}

This is due to the fact that OC's stars are surrounded by other stars with possibly similar
properties. So, setting the number of clusters to two is too low to separate them properly.

As shown in Figure \ref{fig:kmeans_comparisons_melotte_22}, larger values for the number of
clusters allow us to isolate more accurately the resonance in parallax at $\approx 7.3mas$.
However, we have the disadvantage that more groups are formed.

This effect complicates our task of finding the desired open cluster, since we would like
to get just two groups, one for the OC and another with the remaining stars.
Therefore, we have to find a way to set the right value for the number of clusters
to isolate the searched cluster without creating too many groups.

To solve this issue, we will try to estimate the best number of clusters by using the
\emph{silhouette score} \cite{rousseeuw1987silhouettes}.

K-Means does a good job making an initial clustering. However, too many clusters arise
from this characterization and the OC is still polluted with stars that do not belong to it.
Moreover, we would like to reduce the amount of clusters too.

Therefore, we would like to find a better model that improves this initial characterization
by reducing the amount of clusters and also that removes outsiders from the OC.

\subsection{DEC MODEL}

Since we do not have a labeled dataset to train a supervised model,
we have no choice but to use an unsupervised self-trained model.

For that reason we have adapted the \emph{Unsupervised Deep Embedding for Clustering Analysis}
to our work.

The model is composed by a \emph{deep autoencoder}, a \emph{deep encoder} and
a \emph{clustering layer}.

Encoders are used to transform the input data into a latent space
using a non-linear mapping function $f_{\theta} : X \rightarrow Z$.

Although, as explained in section \ref{sec:feature_selection},
the number of features we are managing is not too large,
this latent space helps us reduce the number of features
and avoids the \emph{"curse of dimensionality"} \cite{bellman1961curse}.

These encoders are pretrained before fitting the model to generate predictions. Then,
the encoder is used with the aim of transforming input data to the latent space $Z$. Once the
data has been transformed, a K-Means clusterer is used in order to make an initial clustering.

With that initial configuration, the model iterates alternating between computing an auxiliary
target distribution (Soft Assignment) and minimizing the Kullback-Leiber (KL) divergence
\cite{kullback1951information} to it. This unsupervised algorithm allows us to improve the clustering.

\begin{equation}
  p_{ij} = \frac{q^{2}_{ij} / f_{j}}{\sum_{j'}q^{2}_{ij'}/f_{j'}}
  \label{eq:student_tdistribution}
\end{equation}

In the soft assignment stage,
the \emph{Student's t-distribution} is used as a kernel to measure the similarity
between the embedded points and the cluster centroid.
While in the KL divergence minimization the algorithm iteratively refines clusters by learning
from their high confidence assignments with the help of an auxiliary target distribution.
The model is trained by matching the soft assignment to the target distribution.
The choice of this target distribution is crucial for DEC's performance.
In this work we have taken the target distribution from DEC's original paper \cite{xie2016unsupervised},
which is defined in Equation \ref{eq:student_tdistribution}.

Figure \ref{fig:dec_model_setup} shows the layer setup of our DEC model.
It is simpler than the one tested on the original paper \cite{xie2016unsupervised},
since the number of selected features in our work is smaller than in the original one.
Therefore, using the same configuration would result in a model so
powerful that would incur in overfitting issues unable to make right predictions.

\begin{figure}
  \includegraphics[width=\columnwidth]{../figures/dec_diagram.pdf}
  \caption{DEC model layer setup}
  \label{fig:dec_model_setup}
\end{figure}

Finally, we can refine this selection by filtering those stars which are bellow and above
the 0.10 and 0.90 quantiles for each group, respectively.
That way we remove the most doubtful values from the selection.

All developed resources are available at:
\href{github.com/cdalvaro/machine-learning-master-thesis}{github.com/cdalvaro/machine-learning-master-thesis}
